---
title: "Getting Started with synthdid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with synthdid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
ragg_png = function(..., res = 192) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(dev = "ragg_png", fig.ext = "png")
```

```{r setup, message=FALSE}
library(synthdid)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(12345)
```

# What Is Synthetic Difference-in-Differences?

Panel data---repeated observations of multiple units over time---are the
workhorse of causal inference in the social sciences. When some units receive a
treatment and others do not, two classical approaches exist for estimating the
treatment effect:

- **Difference-in-Differences (DID)** compares the change in outcomes before
  and after treatment between treated and control units. It assumes that, absent
  treatment, both groups would have followed *parallel trends*. DID weights all
  control units and all pre-treatment periods equally.

- **Synthetic Control (SC)** constructs a weighted average of control units that
  closely matches the treated unit's pre-treatment outcome trajectory. It does
  not assume parallel trends in the same way as DID, but it also does not
  difference out time effects---the counterfactual must match the treated unit's
  *level*, not just its trend.

**Synthetic Difference-in-Differences (SDID)**, proposed by Arkhangelsky et al.
(2021), combines both ideas. It selects:

1. **Unit weights** (omega, $\omega$): like SC, a weighted average of control units
   that matches the treated unit's pre-treatment trajectory.
2. **Time weights** (lambda, $\lambda$): like DID, but with weights that emphasize
   pre-treatment periods most relevant for the counterfactual, rather than
   weighting all periods equally.

This double-weighting typically yields better pre-treatment fit than DID and
more robustness than SC. The estimator is implemented in this package via a
Frank-Wolfe optimization algorithm.


# Data Format

The package expects **long-format panel data**: one row per unit-period, with
columns identifying the unit, the time period, the outcome, and a binary
treatment indicator.

```{r}
data("california_prop99")
head(california_prop99)
```

The built-in `california_prop99` dataset contains annual cigarette consumption
(`PacksPerCapita`) for US states from 1970 to 2000. California (`treated = 1`)
passed Proposition 99 in 1989, raising the cigarette tax. All other states
serve as controls.


# Basic Estimation

## The `synthdid()` Formula Interface

The primary entry point is `synthdid()`, which uses a formula interface similar
to `lm()`:

```{r}
result <- synthdid(PacksPerCapita ~ treated,
                   data = california_prop99,
                   index = c("State", "Year"))
```

The formula `PacksPerCapita ~ treated` specifies:

- **Left of `~`**: the outcome variable.
- **Right of `~`**: the binary treatment indicator (0/1).

The `index` argument identifies the two panel dimensions:

- First element: the **unit** identifier (`"State"`).
- Second element: the **time** identifier (`"Year"`).

The returned object supports standard R generics:

```{r}
# Point estimate of the average treatment effect
coef(result)

# Full summary (uses jackknife SE by default for speed)
summary(result, fast = TRUE)
```

## Accessing Weights

The SDID estimate is defined by the weights it selects. You can inspect them
directly:

```{r}
weights <- attr(result, "weights")

# Unit weights: one per control state
head(weights$omega, 10)

# Time weights: one per pre-treatment year
head(weights$lambda, 10)
```

Most omega weights will be zero or near-zero---SDID selects a sparse set of
donor units, much like SC. The time weights typically concentrate on periods
close to the treatment onset.


# The Three Estimators

The `method` argument controls which estimator is used. All three are special
cases of the same underlying framework and share the same interface:

```{r}
sdid_est <- synthdid(PacksPerCapita ~ treated, data = california_prop99,
                     index = c("State", "Year"), method = "synthdid")

sc_est   <- synthdid(PacksPerCapita ~ treated, data = california_prop99,
                     index = c("State", "Year"), method = "sc")

did_est  <- synthdid(PacksPerCapita ~ treated, data = california_prop99,
                     index = c("State", "Year"), method = "did")

data.frame(
  Method   = c("SDID", "SC", "DID"),
  Estimate = round(c(coef(sdid_est), coef(sc_est), coef(did_est)), 2)
)
```

## How They Differ

The three estimators differ in what they fix and what they optimize:

| Estimator | Unit weights ($\omega$) | Time weights ($\lambda$) | Intercept adjustment |
|-----------|------------------------|-------------------------|---------------------|
| **DID**   | Uniform (1/N~0~)       | Uniform (1/T~0~)        | Yes                 |
| **SC**    | Optimized              | All zero                | No                  |
| **SDID**  | Optimized              | Optimized               | Yes                 |

- **DID** passes `omega = rep(1/N0, N0)` and `lambda = rep(1/T0, T0)` ---
  equal weights on everything. The treatment effect is the simple
  difference-in-differences.
- **SC** optimizes omega to match the treated unit's pre-treatment trajectory,
  but sets all lambda to zero and disables the intercept. The counterfactual
  must match the treated unit's *level*.
- **SDID** optimizes both omega and lambda, and includes an intercept. This
  allows it to match trends even when levels differ.

You can switch between methods with `update()` without re-specifying the model:

```{r}
# Start from SDID, switch to DID
did_from_update <- update(sdid_est, method = "did")
coef(did_from_update)
```

## Visualizing the Differences

The built-in `plot()` method shows the treated trajectory, the synthetic
control, and the estimated treatment effect. Comparing across methods reveals
how different weighting strategies produce different counterfactuals:

```{r fig.width=7, fig.height=4}
plot(sdid_est, se.method = "placebo")
```

The parallelogram overlay shows how the estimate is constructed. The blue
segment is the change from the weighted pre-treatment average to the
post-treatment average for California. The red segment is the same change for
the synthetic control. The dashed line extends California's pre-treatment level
by the control's change---this is the counterfactual. The black arrow shows the
treatment effect.

For SC, the counterfactual is constructed differently---no differencing is
involved, so the synthetic control must match California's *level*:

```{r fig.width=7, fig.height=4}
plot(sc_est, se.method = "placebo")
```

For DID, all control states receive equal weight and the counterfactual is a
simple average. Notice that the pre-treatment fit is worse---DID does not
optimize weights to match California's trajectory:

```{r fig.width=7, fig.height=4}
plot(did_est, se.method = "placebo")
```

### Comparing Weight Distributions

The difference between methods becomes concrete when you look at the weights
they assign. SDID and SC concentrate weight on a few informative donors, while
DID spreads it uniformly:

```{r fig.width=8, fig.height=4}
omega_comparison <- data.frame(
  unit = seq_along(attr(sdid_est, "weights")$omega),
  SDID = attr(sdid_est, "weights")$omega,
  SC   = attr(sc_est, "weights")$omega,
  DID  = attr(did_est, "weights")$omega
) |>
  tidyr::pivot_longer(-unit, names_to = "method", values_to = "weight") |>
  dplyr::filter(weight > 0.001)

ggplot(omega_comparison, aes(x = reorder(factor(unit), -weight), y = weight)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  facet_wrap(~ method, scales = "free_x") +
  labs(x = "Control Unit Index", y = "Weight (ω)",
       title = "Unit Weight Distributions by Method",
       subtitle = "Only nonzero weights shown") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_blank(), panel.grid.major.x = element_blank())
```


# Predictions and Effect Curves

The `predict()` method provides three types of output:

## Counterfactual: What Would Have Happened Without Treatment

```{r}
# Counterfactual trajectory for California (what would have happened)
counterfactual <- predict(sdid_est, type = "counterfactual")
counterfactual[, (ncol(counterfactual) - 4):ncol(counterfactual)]
```

## Treated: What Actually Happened

```{r}
# Observed outcomes for California
treated <- predict(sdid_est, type = "treated")
treated[, (ncol(treated) - 4):ncol(treated)]
```

## Effect Curve: Period-by-Period Treatment Effects

The average treatment effect (`coef()`) summarizes the post-treatment period
into a single number. But the effect may vary over time. The effect curve
decomposes it into per-period effects:

```{r fig.width=7, fig.height=4}
effect <- predict(sdid_est, type = "effect")

plot(effect, type = "b", pch = 19,
     xlab = "Post-Treatment Year",
     ylab = "Treatment Effect (packs per capita)",
     main = "Effect of Prop 99 on Cigarette Consumption Over Time")
abline(h = coef(sdid_est), lty = 2, col = "steelblue")
abline(h = 0, col = "grey50")
legend("bottomleft",
       legend = c("Per-period effect", "Average effect"),
       lty = c(1, 2), col = c("black", "steelblue"), bty = "n")
```

The effect grows over time, suggesting that the cigarette tax had a
compounding impact on smoking behavior.


# Standard Errors and Inference

## Computing Standard Errors

Pass `se = TRUE` to compute standard errors during estimation:

```{r}
result_with_se <- synthdid(PacksPerCapita ~ treated,
                           data = california_prop99,
                           index = c("State", "Year"),
                           se = TRUE,
                           se_method = "placebo")

# Standard error is stored as an attribute
attr(result_with_se, "se")
```

## Available SE Methods

Three standard error methods are available:

| Method      | Best for                         | Speed    |
|-------------|----------------------------------|----------|
| `bootstrap` | Multiple treated units           | Slow     |
| `jackknife` | Quick diagnostics                | Moderate |
| `placebo`   | Single treated unit, many controls | Slow  |

**Placebo** re-estimates the treatment effect using each control unit as if it
were treated. The SE is the standard deviation of these placebo estimates. This
is the recommended method when there is a single treated unit and many controls.

**Bootstrap** resamples control units with replacement. It works best with
multiple treated units.

**Jackknife** leaves out one control unit at a time. It is the fastest method
but may underestimate uncertainty.

## Confidence Intervals

```{r}
confint(result_with_se, level = 0.95)
```

## Parallel Processing

Bootstrap and placebo SEs are embarrassingly parallel---each replication is
independent. You can speed them up with the `future` package:

```{r eval=FALSE}
library(future)
plan(multisession, workers = 4)

result <- synthdid(PacksPerCapita ~ treated,
                   data = california_prop99,
                   index = c("State", "Year"),
                   se = TRUE, se_method = "placebo")

plan(sequential)  # Reset when done
```


# Built-In Plotting

## Trajectory Plot

```{r fig.width=7, fig.height=4}
# Default plot: treated vs synthetic control with effect diagram
plot(result_with_se)
```

## Overlaying Trajectories

When trajectories are far apart, it can be hard to assess parallel trends.
Use `overlay` to shift the control toward the treated unit:

```{r fig.width=7, fig.height=4}
# Overlay = 1 aligns pre-treatment levels
plot(result_with_se, overlay = 1)
```

## Unit Contributions

The `synthdid_units_plot()` shows how each control unit contributes to the
estimate. Dot size reflects the unit weight $\omega_i$:

```{r fig.width=7, fig.height=4}
synthdid_units_plot(result_with_se)
```

## Multi-Estimator Comparison

Compare estimators side-by-side using `synthdid_plot()` with a named list:

```{r fig.width=10, fig.height=4}
estimates <- list(
  "Diff-in-Diff"  = did_est,
  "Synth Control"  = sc_est,
  "Synth DID"      = sdid_est
)
synthdid_plot(estimates, se.method = "placebo")
```


# Convergence and Optimization Parameters

The SDID weights are estimated by a Frank-Wolfe (conditional gradient)
optimization algorithm. Like all iterative algorithms, convergence is not
guaranteed within a fixed number of iterations. The package monitors
convergence automatically and warns you when it fails.

## Checking Convergence

```{r}
# Boolean: did the optimization converge?
synthdid_converged(result)

# Detailed diagnostics
synthdid_convergence_info(result)
```

If `synthdid_converged()` returns `FALSE`, the estimate may be unreliable. The
most common fix is to increase `max.iter`.

## Optimization Parameters

The following parameters can be passed via `...` to `synthdid()` to control the
optimization. These are forwarded to `synthdid_estimate()`:

### Regularization

| Parameter            | Default                                  | Description |
|----------------------|------------------------------------------|-------------|
| `eta.omega`          | $(N_1 \cdot T_1)^{1/4}$                 | Regularization strength for unit weights. Higher values shrink omega toward uniform weights. |
| `eta.lambda`         | `1e-6` (near-zero)                       | Regularization strength for time weights. Kept small so lambda is data-driven. |
| `noise.level`        | SD of first-differences of control outcomes | Scale factor that converts eta to zeta: `zeta = eta * noise.level`. |

The regularization penalty in the Frank-Wolfe objective is
$\zeta^2 \cdot \| \omega \|_2^2$ (and analogously for lambda). Larger
`eta.omega` pulls the solution toward uniform weights, making SDID behave
more like DID. Setting `eta.omega` to near-zero makes it behave more like SC.

```{r}
# More regularization: pulls weights toward uniform (DID-like)
result_strong_reg <- synthdid(PacksPerCapita ~ treated,
                              data = california_prop99,
                              index = c("State", "Year"),
                              eta.omega = 10)
coef(result_strong_reg)

# Less regularization: sparser weights (SC-like)
result_weak_reg <- synthdid(PacksPerCapita ~ treated,
                            data = california_prop99,
                            index = c("State", "Year"),
                            eta.omega = 0.01)
coef(result_weak_reg)
```

### Convergence Control

| Parameter         | Default   | Description |
|-------------------|-----------|-------------|
| `max.iter`        | `10000`   | Maximum Frank-Wolfe iterations. Increase if you see convergence warnings. |
| `min.decrease`    | `1e-5 * noise.level` | Stop when the objective decrease per iteration falls below `min.decrease^2`. Smaller values demand tighter convergence. |

```{r}
# Tighter convergence: more iterations, stricter stopping
result_tight <- synthdid(PacksPerCapita ~ treated,
                         data = california_prop99,
                         index = c("State", "Year"),
                         max.iter = 50000,
                         min.decrease = 1e-8)
synthdid_converged(result_tight)
```

### Intercepts

| Parameter           | Default | Description |
|---------------------|---------|-------------|
| `omega.intercept`   | `TRUE`  | Include an intercept when fitting unit weights. Allows the synthetic control to match trends rather than levels. |
| `lambda.intercept`  | `TRUE`  | Include an intercept when fitting time weights. |

Disabling intercepts makes SDID behave more like SC (requiring level matching
rather than trend matching):

```{r}
result_no_intercept <- synthdid(PacksPerCapita ~ treated,
                                data = california_prop99,
                                index = c("State", "Year"),
                                omega.intercept = FALSE)
coef(result_no_intercept)
```

### Sparsification

| Parameter                 | Default            | Description |
|---------------------------|--------------------|-------------|
| `sparsify`                | Internal function  | After an initial optimization, weights below `max(weight)/4` are set to zero and the optimization is re-run. Set to `NULL` to disable. |
| `max.iter.pre.sparsify`   | `100`              | Maximum iterations for the first (pre-sparsification) optimization round. |

Sparsification produces cleaner, more interpretable weights at the cost of
slightly less optimal fit. Disable it if you want the raw Frank-Wolfe solution:

```{r}
result_no_sparse <- synthdid(PacksPerCapita ~ treated,
                             data = california_prop99,
                             index = c("State", "Year"),
                             sparsify = NULL)

# Compare number of nonzero weights
cat("With sparsification:",
    sum(attr(result, "weights")$omega > 1e-6), "nonzero donors\n")
cat("Without sparsification:",
    sum(attr(result_no_sparse, "weights")$omega > 1e-6), "nonzero donors\n")
```

### Pre-Specified Weights

You can supply your own weights instead of estimating them. By default,
supplying a weight vector fixes it (it is not re-optimized). Use
`update.omega = TRUE` or `update.lambda = TRUE` to use supplied weights as
initialization for further optimization instead.

| Parameter        | Default                             | Description |
|------------------|-------------------------------------|-------------|
| `weights`        | `list(omega = NULL, lambda = NULL)` | Pre-specified weight vectors. If non-NULL, used instead of optimizing. |
| `update.omega`   | `FALSE` if `weights$omega` given    | If `TRUE`, use supplied omega as initialization and re-optimize. |
| `update.lambda`  | `FALSE` if `weights$lambda` given   | If `TRUE`, use supplied lambda as initialization and re-optimize. |

```{r}
setup <- attr(result, "setup")

# Use uniform lambda (makes SDID behave like SC + intercept)
result_uniform_lambda <- synthdid(
  PacksPerCapita ~ treated,
  data = california_prop99,
  index = c("State", "Year"),
  weights = list(lambda = rep(1 / setup$T0, setup$T0), omega = NULL)
)
coef(result_uniform_lambda)
```


# Residuals and Diagnostics

The formula interface provides residual extraction for model diagnostics:

```{r}
# Control-unit residuals (how well does the model fit control outcomes?)
resid_control <- residuals(result, type = "control")
cat("Control residuals: ", nrow(resid_control), "units x",
    ncol(resid_control), "periods\n")
cat("Mean absolute residual:", round(mean(abs(resid_control)), 2), "\n")

# Pre-treatment residuals (quality of pre-treatment fit)
resid_pre <- residuals(result, type = "pretreatment")
cat("Pre-treatment RMSE:", round(sqrt(mean(resid_pre^2)), 2), "\n")
```


# Covariate Adjustment

If your data includes time-varying covariates, include them with the `|`
operator in the formula:

```{r eval=FALSE}
# Syntax for covariate adjustment (requires covariates in your data)
result_cov <- synthdid(outcome ~ treatment | covariate1 + covariate2,
                       data = your_data,
                       index = c("unit", "time"))
```

The covariates are used to partial out their effect before estimating the
treatment effect. Internally, a regression coefficient beta is estimated for
each covariate and subtracted from the outcome matrix.


# Summary

## Quick Reference

| Task                    | Code                                                 |
|-------------------------|------------------------------------------------------|
| Estimate SDID           | `synthdid(y ~ w, data, index)`                       |
| Estimate SC             | `synthdid(y ~ w, data, index, method = "sc")`        |
| Estimate DID            | `synthdid(y ~ w, data, index, method = "did")`       |
| Point estimate          | `coef(result)`                                       |
| Confidence interval     | `confint(result)`                                    |
| Summary with SE         | `summary(result)`                                    |
| Effect curve            | `predict(result, type = "effect")`                   |
| Counterfactual          | `predict(result, type = "counterfactual")`            |
| Treated trajectory      | `predict(result, type = "treated")`                   |
| Residuals               | `residuals(result, type = "control")`                 |
| Convergence check       | `synthdid_converged(result)`                          |
| Plot                    | `plot(result)`                                       |
| Unit contributions      | `synthdid_units_plot(result)`                         |
| Switch method           | `update(result, method = "did")`                     |
| Inspect weights         | `attr(result, "weights")`                            |

## Next Steps

- `vignette("formula-interface")` --- deep dive on the formula interface
- `vignette("more-plotting")` --- plotting gallery with customization examples
- `vignette("convergence-diagnostics")` --- troubleshooting convergence
- `vignette("parallel-processing")` --- speeding up SE computation


# References

Arkhangelsky, D., Athey, S., Hirshberg, D. A., Imbens, G. W., & Wager, S.
(2021). Synthetic Difference in Differences. *American Economic Review*,
111(12), 4088--4118.
