---
title: "Plotting Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
library(synthdid)
library(ggplot2)
```

```{r echo=FALSE}
ragg_png <- function(..., res = 192) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(dev = "ragg_png", fig.ext = "png",
                      fig.width = 8, fig.height = 5,
                      warning = FALSE, message = FALSE)
```

The `plot()` method for synthdid objects produces publication-ready
ggplot2 graphics. This vignette walks through every plot type and subtype
available via the `type`, `subtype`, and `show_gap` arguments.

# Setup

We use the California Proposition 99 dataset throughout.

```{r}
data(california_prop99)

result <- synthdid(PacksPerCapita ~ treated,
                   data = california_prop99,
                   index = c("State", "Year"),
                   se = TRUE, se_method = "placebo")
```

# Trajectory Plots

## Default: Treated vs Synthetic Control

The default plot shows the treated and synthetic control trajectories,
treatment onset, a lambda-weight ribbon along the bottom, and the ATT
annotation.

```{r fig.height=4}
plot(result)
```

## With Individual Control Trajectories

Pass `include_controls = TRUE` to draw individual control unit
trajectories behind the main lines. For panels with many controls,
auto-mode selects the top-weighted donors to keep the plot readable.

```{r fig.height=4}
plot(result, include_controls = TRUE)
```

## With Gap Sub-Panel

Pass `show_gap = TRUE` to add a second panel below the trajectories
showing the gap (treated minus synthetic) over time. The post-treatment
area is shaded. This makes it easy to see both the trajectories and the
evolving treatment effect in one figure.

```{r fig.height=6}
plot(result, show_gap = TRUE)
```

## Combining Controls and Gap

```{r fig.height=6}
plot(result, include_controls = TRUE, show_gap = TRUE)
```

# Weight Diagnostics

## Unit Weights (omega)

The lollipop chart shows the largest donor weights. An "Other" bucket
summarizes the rest.

```{r fig.height=4}
plot(result, type = "weights", subtype = "omega")
```

## Time Weights (lambda)

Shows which pre-treatment periods matter most for the synthetic control.
The dashed line marks equal weighting for reference. SDID typically
up-weights periods close to the treatment onset.

```{r fig.height=4}
plot(result, type = "weights", subtype = "lambda")
```

## Cumulative Weight Curve

A Lorenz-style curve showing how many donors account for a given share
of total weight. The orange dot marks the number of units needed to
reach 90% of the total weight.

```{r fig.height=4}
plot(result, type = "weights", subtype = "cumulative")
```

## Combined: Omega + Cumulative

The default `subtype = "both"` places the lollipop and cumulative curve
side by side (requires `patchwork`).

```{r fig.height=4}
plot(result, type = "weights")
```

# Effect Plots

The effect plot shows the per-period treatment effect in the
post-treatment window. The dotted line marks the aggregate ATT. When
standard errors are available, a confidence interval is shown at the
right edge.

```{r fig.height=4}
plot(result, type = "effect")
```

# Diagnostic Plots

## Pre-Treatment Fit

Overlays the treated and synthetic trajectories in the pre-treatment
window with a ribbon showing their gap. The lower panel shows residuals
as a percentage of the treated value. MAPE and RMSE are reported in the
subtitle.
```{r fig.height=5}
plot(result, type = "diagnostic", subtype = "fit")
```

## Convergence Trace

Plots the RMSE of the Frank-Wolfe objective across iterations on a log
scale. Useful for checking that the solver reached a stable solution.

```{r fig.height=4}
plot(result, type = "diagnostic", subtype = "convergence")
```

## Combined Diagnostics

Pass `subtype = "both"` (the default) to stack convergence and fit
panels.

```{r fig.height=6}
plot(result, type = "diagnostic")
```

# Customizing Plots

All plot methods return ggplot2 objects, so you can add layers, change
themes, or modify labels:

```{r fig.height=4}
plot(result) +
  ggtitle("California Proposition 99") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```


# Staggered Adoption Plots

For staggered-adoption estimates, the plot engine supports the same
`type` argument with staggered-specific defaults.

```{r}
# Simulate a small staggered panel
set.seed(42)
N_ctrl <- 20; N_per <- 5; TT <- 20; tau <- 2
cohorts <- c(11, 15)
N <- N_ctrl + length(cohorts) * N_per
Y <- outer(1:N, 1:TT, function(i, t) sin(i / 5) + t / 10) +
  matrix(rnorm(N * TT, sd = 0.5), N, TT)
W <- matrix(0L, N, TT)
idx <- N_ctrl
for (g in cohorts) {
  units <- (idx + 1):(idx + N_per)
  W[units, g:TT] <- 1L
  Y[units, g:TT] <- Y[units, g:TT] + tau
  idx <- idx + N_per
}
rownames(Y) <- rownames(W) <- paste0("u", 1:N)
colnames(Y) <- colnames(W) <- 1:TT

stag_df <- data.frame(
  unit = rep(rownames(Y), each = TT),
  time = rep(as.integer(colnames(Y)), N),
  outcome = c(t(Y)),
  treatment = c(t(W)),
  stringsAsFactors = FALSE
)

stag_result <- synthdid(outcome ~ treatment, data = stag_df,
                         index = c("unit", "time"),
                         adoption = "staggered")
```

## Cohort Effects (default)

Bar chart of per-cohort treatment effects with the aggregate ATT line.

```{r fig.height=4}
plot(stag_result)
```

## Event Study

Effect paths by cohort in event time (periods since treatment).

```{r fig.height=4}
plot(stag_result, type = "effect", subtype = "event")
```

## Per-Cohort Trajectories

Treated vs synthetic for each cohort, arranged via `patchwork`.

```{r fig.height=5}
plot(stag_result, type = "trajectory")
```

## Aggregation Weights

How much each cohort contributes to the aggregate ATT.

```{r fig.height=4}
plot(stag_result, type = "weights")
```


# Legacy API

The classic `synthdid_plot()` and related functions still work. They are
called automatically when legacy arguments (e.g., `spaghetti.units`,
`facet`, `overlay`) are passed to `plot()`.

```{r fig.height=4}
# Spaghetti plot via legacy argument
top_controls <- synthdid_controls(result)[1:10, , drop = FALSE]
plot(result, spaghetti.units = rownames(top_controls))
```

For multi-estimator comparisons, use `synthdid_plot()` directly:

```{r fig.width=10, fig.height=4}
sc_result  <- update(result, method = "sc")
did_result <- update(result, method = "did")

estimates <- list(
  "Diff-in-Diff"  = did_result,
  "Synth Control" = sc_result,
  "Synth DID"     = result
)
synthdid_plot(estimates, se.method = "placebo")
```
