---
title: "Staggered Adoption"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Staggered Adoption}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
ragg_png = function(..., res = 192) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(dev = "ragg_png", fig.ext = "png")
```

```{r setup, message=FALSE}
library(synthdid)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(42)
```

# Introduction

The standard SDID estimator from Arkhangelsky et al. (2021) assumes that all
treated units adopt treatment simultaneously. In practice, many policies are
adopted at different times across units---states pass laws in different years,
firms adopt technologies in different quarters, and so on. This is called
**staggered adoption**.

Porreca (2022, *Economics Letters*) formalizes an extension of SDID to the
staggered setting, building on the idea sketched in Appendix A of
Arkhangelsky et al. (2021). The approach decomposes the staggered panel into
cohort-specific subproblems, estimates SDID on each, and aggregates the
results into a single average treatment effect on the treated (ATT).

This vignette replicates the empirical application from Porreca (2022) using
data on sentencing enhancement laws from Abrams (2012), then demonstrates
additional features of the staggered estimator.


# The Sentencing Enhancement Data

Abrams (2012) studies whether sentencing enhancements for firearm offenses
deter armed robbery. Twenty-three US states adopted "add-on" gun laws at
different times between 1972 and 1996, while 17 states never adopted such
laws. The `abrams_sentencing` dataset is a balanced panel of 40 states
observed annually from 1965 to 2002.

```{r}
data(abrams_sentencing)
head(abrams_sentencing)
```

The staggered adoption structure is rich: 14 distinct adoption cohorts spread
over 25 years.

```{r}
abrams_sentencing |>
  filter(!is.na(addonyr)) |>
  distinct(state, addonyr) |>
  count(addonyr, name = "n_states")
```

We can visualize the treatment pattern:

```{r fig.width=7, fig.height=5}
abrams_sentencing |>
  mutate(
    status = case_when(
      is.na(addonyr) ~ "Never treated",
      post == 1 ~ "Treated",
      TRUE ~ "Not yet treated"
    ),
    status = factor(status, levels = c("Never treated", "Not yet treated", "Treated"))
  ) |>
  ggplot(aes(x = year,
             y = reorder(factor(state),
                         -as.numeric(is.na(addonyr)) * 10000 - ifelse(is.na(addonyr), 0, addonyr)),
             fill = status)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_manual(values = c("Never treated" = "grey85",
                                "Not yet treated" = "steelblue",
                                "Treated" = "firebrick")) +
  labs(x = "Year", y = "State", fill = NULL,
       title = "Treatment Adoption Pattern",
       subtitle = "23 states adopted sentencing enhancements between 1972 and 1996") +
  theme_minimal(base_size = 10) +
  theme(axis.text.y = element_blank(),
        panel.grid = element_blank())
```


# The Problem with Naive SDID

Why can't we just apply the standard (simultaneous) SDID estimator to a
staggered panel? The core issue is **control group contamination**: when
units adopt treatment at different times, already-treated units end up in the
control group for later-adopting cohorts. This violates a fundamental
assumption of the synthetic control approach---that control units are
untreated throughout the study period.

The `synthdid()` function detects staggered adoption and raises an error if
you try to force simultaneous estimation:

```{r error=TRUE}
synthdid(pcrrobgun ~ post,
         data = abrams_sentencing,
         index = c("state", "year"),
         adoption = "simultaneous")
```


# Staggered SDID Methodology

The Porreca (2022) approach handles staggered adoption by decomposing the
problem into cohort-specific **2x2 subproblems**:

1. **Identify cohorts**: group treated units by their adoption time. In the
   Abrams data, this produces 14 cohorts.

2. **Construct subproblems**: for each cohort, build a standard SDID problem
   with:
   - **Treated units**: the cohort members
   - **Control units**: determined by `control_type`
   - **Pre-treatment periods**: all periods before the cohort's adoption
   - **Post-treatment periods**: periods after adoption

3. **Estimate**: run standard SDID on each subproblem independently.

4. **Aggregate**: combine cohort-level effects into a single ATT using
   aggregation weights.

Two key choices govern the construction:

**Control set** (`control_type`):

- `"not_yet_treated"` (default): controls include both never-treated units
  and units that haven't adopted yet at the time of the cohort's last
  post-treatment period. This maximizes the donor pool.
- `"never_treated"`: only units that never adopt serve as controls.

**Aggregation weights** (`weights`):

- `"treated_unit_time"` (default): each cohort is weighted by the number of
  treated unit-time cells (cohort size x post-treatment periods). This gives
  more weight to cohorts that affect more units for longer.
- `"cohort_size"`: weights by the number of treated units only.
- `"equal"`: each cohort gets equal weight regardless of size.


# Replicating Porreca (2022)

Porreca (2022, Table 1) reports a SynthDiD estimate of $-16.697$ for the
effect of sentencing enhancements on per-capita armed robbery rates, using
never-treated controls. We can replicate this exactly:

```{r}
result <- synthdid(pcrrobgun ~ post,
                   data = abrams_sentencing,
                   index = c("state", "year"),
                   control_type = "never_treated",
                   weights = "treated_unit_time")

result
```

```{r}
coef(result)
```

The aggregate ATT of $-16.697$ matches Porreca's Table 1 exactly. The
estimate suggests that sentencing enhancements reduced per-capita armed
robbery by about 17 per 100,000 population.


## Cohort-Level Effects

The aggregate hides substantial heterogeneity across cohorts:

```{r}
coef(result, type = "cohort")
```

The full cohort table shows each cohort's estimate alongside its size and
aggregation weight:

```{r}
result$cohort_effects
```


# Comparing Methods: SDID vs DID vs SC

All three estimation methods work with staggered adoption. We can compare
them on the same data:

```{r}
did_result <- synthdid(pcrrobgun ~ post, data = abrams_sentencing,
                       index = c("state", "year"), method = "did",
                       control_type = "never_treated",
                       weights = "treated_unit_time")

sc_result <- synthdid(pcrrobgun ~ post, data = abrams_sentencing,
                      index = c("state", "year"), method = "sc",
                      control_type = "never_treated",
                      weights = "treated_unit_time")

comparison <- data.frame(
  Method = c("SDID", "DID", "SC"),
  ATT = round(c(coef(result), coef(did_result), coef(sc_result)), 3),
  Porreca_Table1 = c(-16.697, NA, NA)
)
comparison
```

Porreca's Table 1 also reports TWFE = $-20.182$ and Callaway & Sant'Anna =
$-23.724$. The SynthDiD estimate is more conservative than these
alternatives, consistent with Porreca's finding that SDID may be more
appropriate when the data-generating process follows a latent factor model
rather than additive fixed effects.


# Visualization

## Cohort Effects

The `plot()` method produces a cohort-level bar plot. Bar opacity reflects the
aggregation weight, and the dashed red line marks the aggregate ATT:

```{r fig.width=7, fig.height=4}
plot(result)
```

## Event Study

An event-study style plot shows how treatment effects evolve over time since
adoption, separately for each cohort:

```{r fig.width=7, fig.height=4}
plot(result, type = "event")
```

The event-study view reveals whether effects are immediate or delayed, and
whether they grow, shrink, or remain stable after adoption.


# Sensitivity Analysis

## Aggregation Weights

The choice of aggregation weights determines how cohort-level effects are
combined:

```{r}
result_cohort_size <- synthdid(pcrrobgun ~ post, data = abrams_sentencing,
                               index = c("state", "year"),
                               control_type = "never_treated",
                               weights = "cohort_size")

result_equal <- synthdid(pcrrobgun ~ post, data = abrams_sentencing,
                         index = c("state", "year"),
                         control_type = "never_treated",
                         weights = "equal")

data.frame(
  aggregation = c("treated_unit_time", "cohort_size", "equal"),
  ATT = round(c(coef(result), coef(result_cohort_size), coef(result_equal)), 3)
)
```

## Control Set: Not-Yet-Treated vs Never-Treated

The default `control_type = "not_yet_treated"` uses a larger donor pool. With
many cohorts spread over time (as in this dataset), not-yet-treated units
substantially expand the control set for early adopters:

```{r}
result_nyt <- synthdid(pcrrobgun ~ post, data = abrams_sentencing,
                       index = c("state", "year"),
                       control_type = "not_yet_treated",
                       weights = "treated_unit_time")

data.frame(
  control_set = c("never_treated", "not_yet_treated"),
  ATT = round(c(coef(result), coef(result_nyt)), 3)
)
```


# Inference

Bootstrap standard errors are available via `se = TRUE`. The bootstrap
resamples entire units while preserving the cohort structure.

```{r eval=FALSE}
result_se <- synthdid(pcrrobgun ~ post,
                      data = abrams_sentencing,
                      index = c("state", "year"),
                      control_type = "never_treated",
                      weights = "treated_unit_time",
                      se = TRUE,
                      se_method = "bootstrap",
                      se_replications = 200)

summary(result_se)
confint(result_se)
```

For faster computation, the bootstrap is parallelizable using the `future`
framework. See `vignette("parallel-processing")` for details:

```{r eval=FALSE}
library(future)
plan(multisession, workers = 4)

result_se <- synthdid(pcrrobgun ~ post,
                      data = abrams_sentencing,
                      index = c("state", "year"),
                      control_type = "never_treated",
                      weights = "treated_unit_time",
                      se = TRUE)

plan(sequential)
```


# References

Abrams, D. S. (2012). Estimating the Deterrent Effect of Incarceration using
Sentencing Enhancements. *American Economic Journal: Applied Economics*,
4(4), 32--56.

Arkhangelsky, D., Athey, S., Hirshberg, D. A., Imbens, G. W., & Wager, S.
(2021). Synthetic Difference in Differences. *American Economic Review*,
111(12), 4088--4118.

Porreca, M. (2022). Synthetic Difference in Differences Estimation with
Staggered Treatment Timing. *Economics Letters*, 220, 110873.
